---
title: "BCW"
author: "Aine"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading Libraries: 
```{r, include=FALSE}
library(BART)
library(dcurves)
library(randomForest)
library(gtsummary)
library(broom.helpers)
library(e1071)
library(caret) 
library(ggcorrplot)
library(MASS) #For descrimminate analysis
library(pROC) #ROC curve
library(dplyr) 
library(knitr)
library(kableExtra)
library(ggplot2)
library(glmnet)#Lasso/ Ridge
library(gam)
library(boot)
library(pls)
library(gbm)
library(class)
library(CalibrationCurves) 
library(corrplot)
```

Loading Dataset:
```{r}
#Breast Cancer Wisconsin (Diagnostic)
BCWdata <- read.csv("C:wisconsin_dbc.csv")
```
Review Dataset: 
```{r}
BCW <- BCWdata %>%
  dplyr::select(-"ID") %>%                                    # removing the ID
  mutate(malignant = ifelse(Diagnosis == "M", 1, 0)) %>%  #numeric
  dplyr::select(-"Diagnosis")                                 
summary(BCW)
cor_mat <- cor(BCW[sapply(BCW, is.numeric)], use="pairwise.complete.obs")
cor_mat <- as.matrix(cor_mat)
ggcorrplot(cor_mat)

correlation = as.data.frame(as.table(cor_mat)) %>%
  mutate(Var1 = as.character(Var1), Var2 = as.character(Var2)) %>% #lets filter duplicates out 
  filter(Var1 < Var2) %>% #removes duplicates
  filter(abs(Freq) > 0.7) %>% 
  arrange(desc(abs(Freq))) #decending order
kable(correlation, digits = 2, caption = "Correlations greater than .7")

total_correlation <- nrow(correlation)
total_correlation
```

Split into Test/ Train:
```{r}
set.seed(321321) #reproducible
n<-nrow(BCW)
train.BCW<-runif(n)<0.7
BCW.test<-BCW[!train.BCW,]
V2.test<-BCW.test$malignant
V2.test.factor <- as.factor(BCW.test$malignant)
```

Create Reproducible Brier Score: 
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}
```


## Logistic Regression: 
Brier's Score: 
```{r}
glm.BCW<-glm(malignant ~., data = BCW, subset = train.BCW, family = binomial)
BCW.probs<-predict(glm.BCW, newdata = BCW.test, type = "response") 
GLM_Brier = brier_score(V2.test, BCW.probs)
GLM_Brier
```
Confusion Matrix & Accuracy Score: 

Threshold: A threshold of .3 was chosen to ensure high sensitivity with respect to specify. A threshold of .3 helps reduce the risk of false negatives which would have devastating consequences with respect to classifying tumors.  

Confusion Matrix & Accuracy:
```{r}
glm.pred <- rep(0, nrow(BCW.test)) #size of the test set.
glm.pred[BCW.probs > 0.3] = 1#.3threshold for classification
table(glm.pred, V2.test)
GLM.Accuracy = mean(glm.pred == V2.test)
GLM.Accuracy
```
Roc Curve: 
```{r}
roc_obj.glm <- roc(V2.test, BCW.probs)
plot(roc_obj.glm, type = "s")
GLM.AUC = auc(roc_obj.glm)
GLM.AUC
```
Calibration CUrve: 
```{r}
valProbggplot(BCW.probs, V2.test)
```

## Stepwise Appraoches: 
Using stepwise approach and choosing 'both' allowed the model to evaluate predictors using both forward and backward steps. It's final model includes only 13 predictors. It's classification ability is evaluated below: 
```{r, warning=FALSE}
set.seed(123)
glm.step <- step(glm.BCW, direction = "both", trace = FALSE) # Stepwise (both forward & backward) selection based on AIC. FALSE= dont see how got to choosen values. 
summary(glm.step)
```
Brier Score:
```{r}
step.probs<-predict(glm.step, newdata = BCW.test, type = "response") 
step_Brier = brier_score(V2.test, step.probs) 
step_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.step.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.step.pred[step.probs > 0.3] = 1 #.5threshold for classification
table(BCW.step.pred, V2.test)
Accuracy.step = mean(BCW.step.pred == V2.test)
Accuracy.step
```

Roc Curve: 
```{r}
roc_obj.step <- roc(V2.test, step.probs)
plot(roc_obj.step, type = "s")
step.AUC = auc(roc_obj.step)
step.AUC
```
Calibration CUrve: 
```{r}
valProbggplot(step.probs, V2.test)
```


```{r}
STEP <- data.frame(Model = c("Brier Score", 'Accuracy Score', 'Area Under the Curve'), STEP = c(step_Brier,Accuracy.step,step.AUC))

STEP %>%
  kable(caption = "Stepwise Subset Selection", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

## Ridge Regression: 
```{r}
set.seed(123)
x <- model.matrix(malignant ~ ., BCW)[, -1] #make matrixes
y <- BCW$malignant
ridge.mod<-glmnet(x[train.BCW,], y[train.BCW], alpha = 0, family = binomial)
cv.out <- cv.glmnet(x[train.BCW, ], y[train.BCW], alpha = 0, family = binomial)
plot(cv.out)
bestlam <- cv.out$lambda.min #May prefer 1-SE not minimum but here minimum. 
bestlam
#cv.out$lambda.1se?
```

Brier Score:
```{r}
ridge.probs <- predict(ridge.mod, s = bestlam,newx = x[!train.BCW, ], type = "response")
Ridge_Brier = brier_score(V2.test, ridge.probs)
Ridge_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.ridge.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.ridge.pred[ridge.probs > .3] <- 1 #.3threshold for classification
ridge.probs <- as.vector(ridge.probs)
table(BCW.ridge.pred, V2.test)
Accuracy.Ridge = mean(BCW.ridge.pred == V2.test)
Accuracy.Ridge
```
ROC Curve:
```{r}

roc_obj.ridge <- roc(V2.test, ridge.probs)
plot(roc_obj.ridge, type = "s")
Ridge.AUC = auc(roc_obj.ridge)
Ridge.AUC
```
Calibration Plot: 
```{r}
Ridge.calplot<-valProbggplot(ridge.probs, V2.test)
Ridge.calplot
```
## Lasso Model: 
```{r}
set.seed(321321)
lasso.mod<-glmnet(x[train.BCW,], y[train.BCW], alpha = 1, family = binomial)
cv.out <- cv.glmnet(x[train.BCW, ], y[train.BCW], alpha = 1, family = binomial)
plot(cv.out)
bestlam <- cv.out$lambda.min #May prefer 1-SE not minimum but here minimum.
bestlam
#cv.out$lambda.1se
Kept <- coef(cv.out, s = "lambda.min")
Kept[Kept[,1] != 0, ]
```
Brier's Score:
```{r}
lasso.probs <- predict(lasso.mod, s = bestlam,newx = x[!train.BCW, ], type = "response")
lasso_Brier = brier_score(V2.test, ridge.probs)#Ensure factors are 1 & 0. 
lasso_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.lasso.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.lasso.pred[lasso.probs > .3] <- 1 #.3threshold for classification
table(BCW.lasso.pred, V2.test)
Accuracy.lasso = mean(BCW.lasso.pred == V2.test)
Accuracy.lasso
```
Roc Curve:
```{r}
roc_obj.lasso <- roc(V2.test, ridge.probs)
plot(roc_obj.lasso, type = "s")
lasso.AUC = auc(roc_obj.lasso)
lasso.AUC
```
Calibration Plot: 
```{r}
Lasso.calplot<-valProbggplot(lasso.probs, V2.test)
Lasso.calplot
```
## Matrix Test/ Train: 
```{r}
set.seed(123)
standardized.X <- scale(BCW[, names(BCW) != "malignant"]) #standardize & sure malignant isnt 
Y <- as.factor(BCW$malignant)#y must be a factor in PCA
X <- standardized.X[train.BCW, ]
XTest <- standardized.X[!train.BCW, ]
Ytrain <- Y[train.BCW]
Ytest <- Y[!train.BCW]
Ytest_numeric = as.numeric(as.character(Ytest)) #going factor-numeric causes it to fail
Ytrain_numeric = as.numeric(as.character(Ytrain))
```

## PCA Approach: 
```{r}
plsda.fit <- plsda(X, Ytrain, cv = TRUE)
validationplot(plsda.fit, val.type = "MSEP")
```
From this plot we can wee that having one PCA component preforms best on the training data. 
Brier's Score
```{r}
plsda.fit <- plsda(X, Ytrain, ncomp = 1)
plsda.probs <- predict(plsda.fit, newdata = XTest, type = "prob")

prob_M <- plsda.probs[, "1", 1]
Ytest_numeric <- as.numeric(as.character(Ytest))#becuase made y a factor back to numeric

plsda_Brier <- brier_score(Ytest_numeric, prob_M)
plsda_Brier
```
Confusion Matrix & Accuracy Score:
```{r}
BCW.plsda.pred <- rep(0, length(prob_M))
BCW.plsda.pred[prob_M > 0.3] <- 1
table(BCW.plsda.pred, Ytest)
Accuracy.plsda = mean(BCW.plsda.pred == Ytest)
Accuracy.plsda
```

Roc Curve: 
```{r}
roc_obj.plsda <- roc(Ytest_numeric, prob_M)
plot(roc_obj.plsda, type = "s")
plsda.AUC = auc(roc_obj.plsda)
plsda.AUC
```
Calibration Plot: 
```{r}
plsda.calplot<-valProbggplot(prob_M, Ytest_numeric)
plsda.calplot
```
## Linear Discriminant Analysis:
```{r}
lda.fit <- lda(malignant ~., data = BCW, subset = train.BCW)
lda.pred <- predict(lda.fit, BCW.test)
lda.class <- lda.pred$class
lda.probs <- lda.pred$posterior
prob_M <- lda.probs[, 2]
```
Brier's Score: 
```{r}
lda_Brier <- brier_score(V2.test, prob_M)
lda_Brier
```

Confusion Matrix & Accuracy Score:
```{r}
BCW.lda.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.lda.pred[prob_M > .3] <- 1 #.5threshold for classification
table(BCW.lda.pred, V2.test)
Accuracy.lda = mean(BCW.lda.pred == V2.test)
Accuracy.lda
```
Assessing Results: 
```{r}
summary(prob_M)  
hist(prob_M)   
```

Roc Curve:
```{r}
roc_obj.lda <- roc(V2.test, prob_M)
plot(roc_obj.lda, type = "s")
lda.AUC = auc(roc_obj.lda)
lda.AUC
```
Calibration Plot: 
```{r}
lda.calplot<-valProbggplot(prob_M, V2.test)
lda.calplot
```
## Naive Bayes
```{r}
nb.fit <- naiveBayes(malignant ~ ., data = BCW, subset = train.BCW) #TEST/ TRAIN SPLIT
```
Brier's Score:
```{r}
nb.pred <- predict(nb.fit, BCW.test, type = "raw")  #again predicted on the test set 
nb.probs <- nb.pred[, 2]#extracts column of probability=malignant 
NB_Brier = brier_score(V2.test, nb.probs)#Ensure factors are 1 & 0. 
NB_Brier
```

Confusion Matrix & Accuracy: 
```{r}
nb.class= rep(0, nrow(BCW.test))  
nb.class[nb.probs > 0.3] = 1  
nb.class <- factor(nb.class, levels = c(0,1))   
table(nb.class, V2.test)
NB.Accuracy <- mean(nb.class == V2.test)
NB.Accuracy
```

Assessing Results: 
```{r}
summary(nb.probs)  
hist(nb.probs)   
```
Roc Curve:
```{r}
roc_obj.NB <- roc(V2.test, nb.probs)
plot(roc_obj.NB, type = "s", col = "blue")

NB.AUC = auc(roc_obj.NB)
NB.AUC
```

Calibration Curve: 
```{r}
valProbggplot(nb.probs, V2.test)
```
## K-NN
```{r}
set.seed(123)
k.values <- 1:20 #k can be 1 to 20
cv.error <- rep(NA, 20) #stores misclassification error
for(i in k.values){
  pred <- knn.cv(train = X,cl = Ytrain,k = i) #leave-one-out CV 
  cv.error[i] <- mean(pred != Ytrain) #stores misclassification erro
}
best.k <- which.min(cv.error)
best.k
```
Brier Score: 
```{r}
knn.fit <- knn(X, XTest, Ytrain, k = 3, prob = TRUE)
knn.probs <- attr(knn.fit, "prob") #knn for classification couldnt use predict? 
prob_M <- ifelse(knn.fit == 1, knn.probs, 1 - knn.probs)
knn_Brier = brier_score(Ytest_numeric, prob_M)
knn_Brier
```
Accuracy & Confusion Matrix:
```{r}
knn.class= rep(0, nrow(BCW.test))  
knn.class[prob_M > 0.3] = 1  
table(knn.class, Ytest_numeric)
knn.Accuracy = mean(knn.class==Ytest_numeric)
knn.Accuracy
```

```{r}
summary(prob_M)
hist(prob_M)
```

Roc Curve:
```{r}
roc_obj.knn<- roc(Ytest, prob_M)
plot(roc_obj.knn, type = 's')
knn.AUC = auc(roc_obj.knn)
knn.AUC
```
Calibration Curve: 
```{r}
knn.calplot<-valProbggplot(prob_M, Ytest_numeric)
knn.calplot 
```

## Random Forest: 
```{r}
#As now will be doing classification tasks so should be a factor. 
rf.fit <- randomForest(as.factor(malignant) ~ ., data = BCW, subset =train.BCW, mtry = 5, importance = TRUE)
varImpPlot(rf.fit)
```
CV for m: 
```{r}
set.seed(321)
mtry.values <- c(2, 5, 8, 10, 15)

mtry.errors <- sapply(mtry.values, function(m){
  fit <- randomForest(as.factor(malignant) ~., data = BCW[train.BCW, ], mtry = m, ntree = 500)
  return(fit$err.rate[500, "OOB"])  
})

best.mtry <- mtry.values[which.min(mtry.errors)]
best.mtry
```
Brier Score: 
```{r}
rf.probs <- predict(rf.fit, newdata = BCW.test, type = "prob")
prob_M <- rf.probs[, 2]
rf_Brier <- brier_score(V2.test, prob_M)
rf_Brier
```
Accuracy & Confusion Matrix:
```{r}
rf.class= rep(0, nrow(BCW.test))  
rf.class[prob_M > 0.3] = 1
rf.class = factor(rf.class, levels = c(0,1))
table(rf.class, V2.test)
rf.Accuracy = mean(rf.class==V2.test)
rf.Accuracy
```
Assessing Predictions: 
```{r}
summary(prob_M)
hist(prob_M)
```


Roc Curve:
```{r}
roc_obj.rf<- roc(V2.test, prob_M)
plot(roc_obj.rf, type = 's')
rf.AUC = auc(roc_obj.rf)
rf.AUC
```
Calibration Curve: 
```{r}
rf.calplot<-valProbggplot(prob_M, V2.test)
rf.calplot 
```
## Boosting:
I trialed different Boosting Models:
with different number of trees, different shrinkage and different interaction depths. 
```{r}
set.seed(321)
boost.fit1000 <- gbm(malignant ~., data = BCW[train.BCW, ],distribution = "bernoulli",n.trees = 1000,interaction.depth = 4,shrinkage = 0.01,cv.folds = 5)

best.trees <- gbm.perf(boost.fit1000, method = "cv")
best.trees
```
Shrinkage:
```{r}
set.seed(321)
shrinkage.values <- c(0.001, 0.01, 0.05, 0.1, 0.2) 

shrinkage.errors <- sapply(shrinkage.values, function(s){
  fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = 585,    interaction.depth = 4,shrinkage = s, cv.folds = 5) 
  return(min(fit$cv.error))       
})

best.shrinkage <- shrinkage.values[which.min(shrinkage.errors)]
best.shrinkage
```
Interaction Depth: 
```{r}
set.seed(321)
id <- c(2, 3, 4, 5) 

depth.errors <- sapply(id, function(d){
  fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = 585,    interaction.depth = d,shrinkage = 0.2, cv.folds = 5)
  return(min(fit$cv.error))
})

best.depth <- id[which.min(depth.errors)]
best.depth
```
Brier Score: 
```{r}
boost.fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = 585,    interaction.depth = 4,shrinkage = 0.02) #optimum fit
summary(boost.fit)

boost.probs <- predict(boost.fit, newdata = BCW.test, type = "response") #vector of probs 
boost_Brier <- brier_score(V2.test, boost.probs)
boost_Brier
```
Accuracy & Confusion Matrix:
```{r}
boost.class= rep(0, nrow(BCW.test))  
boost.class[boost.probs > 0.3] = 1  
boost.class = factor(boost.class, levels = c(0, 1))
table(boost.class, V2.test)
boost.Accuracy = mean(boost.class==V2.test)
boost.Accuracy
```
Assessing Results: 
```{r}
summary(boost.probs)  
hist(boost.probs)   
```


Roc Curve:
```{r}
roc_obj.boost<- roc(V2.test, boost.probs)
plot(roc_obj.boost, type = 's')
boost.AUC = auc(roc_obj.boost)
boost.AUC
```
Calibration Curve: 
```{r}
boost.calplot<-valProbggplot(boost.probs, V2.test)
boost.calplot 
```
## Bayesian Additive Regression Trees
```{r}
bart.fit <- pbart(X, Ytrain_numeric, x.test = XTest)
```
Brier Score: 
```{r}
prob_M <- bart.fit$prob.test.mean
bart_Brier <- brier_score(V2.test, prob_M)
bart_Brier 
```
Accuracy & Confusion Matrix:
```{r}
bart.class <- factor(ifelse(prob_M > 0.5, 1, 0), levels = c(0, 1))
table(bart.class, V2.test)
Accuracy.bart <- mean(bart.class == V2.test.factor)
Accuracy.bart
```
Assessing Results: 
```{r}
summary(prob_M)  
hist(prob_M)   
```


Roc Curve:
```{r}
roc_obj.bart<- roc(V2.test, prob_M)
plot(roc_obj.bart, type = 's')
bart.AUC = auc(roc_obj.bart)
bart.AUC
```
Calibration Curve: 
```{r}
bart.calplot<-valProbggplot(prob_M, V2.test)
bart.calplot 
```

## Conclusion:
Brier Scores: 
```{r}
brier_scores <- data.frame(Model = c("GLM", "Best Subset Model", "Ridge Model", "Lasso Model", 'PCA Model', 'LDA Model', 'Naive Bayes', 'K-NN'), Brier_Score = c(GLM_Brier, step_Brier, Ridge_Brier, lasso_Brier, plsda_Brier, lda_Brier, NB_Brier, knn_Brier)
)

brier_scores %>%
  kable(caption = "Brier Scores for Different Models", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```
Accuracy Scores: 
```{r}
Accuracy_scores <- data.frame(Model = c("GLM", "Best SUbset", "Ridge Model", "Lasso Model", 'PCA Model', 'LDA Model', 'Naive Bayes', 'K-NN'), Accuracy_Score = c(GLM.Accuracy,Accuracy.step, Accuracy.Ridge, Accuracy.lasso, Accuracy.plsda, Accuracy.lda, NB.Accuracy, knn.Accuracy)
)

Accuracy_scores %>%
  kable(caption = "Accuracy Scores for Different Models", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


AUC Scores: 
```{r}
AUC_scores <- data.frame(Model = c("GLM", "Best SUbset", "Ridge Model", "Lasso Model", 'PCA Model', 'LDA Model', 'Naive Bayes', 'K-NN'), AUC_Score = c(GLM.AUC,step.AUC, Ridge.AUC, lasso.AUC, plsda.AUC, lda.AUC, NB.AUC, knn.AUC)
)

AUC_scores %>%
  kable(caption = "AUC Scores for Different Models", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```
Get all on one calibration & roc curves? 
```{r}
plot(roc_obj.glm, type = "s", col = "blue")
lines(roc_obj.step, type = "s", col = "red") #change the names of the rocs to the datas
lines(roc_obj.ridge, type = "s", col = "purple")
lines(roc_obj.lasso, type = "s", col = "green")
lines(roc_obj.plsda, type = "s", col = "yellow")
lines(roc_obj.lda, type = "s", col = "orange")
lines(roc_obj.NB, type = "s", col = "pink")
lines(roc_obj.knn, type = "s", col = "brown")
```

Clinical Usefullness:

```{r}
df_dca <- data.frame(
  V2 = V2.test,
  lasso = as.vector(lasso.probs)
)

dcurves::dca(
  V2 ~ lasso,
  data = df_dca,
  thresholds = seq(0, 0.35, 0.01),
  label = list(BCWpred = "Prediction Model")
)

dca_results <- dcurves::dca(
  V2 ~ lasso,
  data = df_dca,
  as_probability = "lasso",
  thresholds = c(0.05, 0.35, 0.75)
)

dca_results %>%
  as_tibble() %>%
  select(label, threshold, net_benefit) %>%
  gt::gt() %>%
  gt::fmt_percent(columns = threshold, decimals = 0) %>%
  gt::fmt(columns = net_benefit, fns = function(x) style_sigfig(x, digits = 3)) %>%
  gt::cols_label(
    label = "Strategy",
    threshold = "Decision Threshold",
    net_benefit = "Net Benefit"
  ) %>%
  gt::cols_align("left", columns = label)
```

