---
title: "Breast Cancer Wisconsin Dataset"
author: "Aine"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading Libraries: 
```{r, include=FALSE}
library(BART)
library(tree)
library(dcurves)
library(randomForest)
library(gtsummary)
library(broom.helpers)
library(e1071)
library(caret) 
library(ggcorrplot)
library(MASS) #For descrimminate analysis
library(pROC) #ROC curve
library(dplyr) 
library(knitr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(glmnet)#Lasso/ Ridge
library(gam)
library(boot)
library(pls)
library(gbm)
library(dbarts)
library(class)
library(CalibrationCurves) 
library(corrplot)
```

Loading Dataset:
```{r}
#Breast Cancer Wisconsin (Diagnostic)
BCWdata <- read.csv("wisconsin_dbc.csv")
```
Review Dataset: 
```{r}
BCW <- BCWdata %>%
  dplyr::select(-"ID") %>%                                    # removing the ID
  mutate(malignant = ifelse(Diagnosis == "M", 1, 0)) %>%  #numeric
  dplyr::select(-"Diagnosis")                                 
summary(BCW)
cor_mat <- cor(BCW[sapply(BCW, is.numeric)], use="pairwise.complete.obs")
cor_mat <- as.matrix(cor_mat)
ggcorrplot(cor_mat)

correlation = as.data.frame(as.table(cor_mat)) %>%
  mutate(Var1 = as.character(Var1), Var2 = as.character(Var2)) %>% #lets filter duplicates out 
  filter(Var1 < Var2) %>% #removes duplicates
  filter(abs(Freq) > 0.7) %>% 
  arrange(desc(abs(Freq))) #decending order

total_correlation <- nrow(correlation)
total_correlation
```

```{r}
correlation %>%
  head(10) %>%
  kable(digits = 2, caption = "Top 10 Correlations:") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 

  
```
Assessing for Extreme Values: 
```{r}
library(psych) 
desc <- describe(BCW, fast = TRUE)
desc[ , c("min", "max", "mean", "median", "range")]
```


Split into Test/ Train:
```{r}
set.seed(321321) #reproducible
n<-nrow(BCW)
train.BCW<-runif(n)<0.7
BCW.test<-BCW[!train.BCW,]
V2.test<-BCW.test$malignant
V2.test.factor <- as.factor(BCW.test$malignant)

table(BCW$malignant[train.BCW]) %>%
  kable(digits = 2, caption = "Number of Events (training set):") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 

table(BCW$malignant[!train.BCW])%>%
  kable(digits = 2, caption = "Number of Events (test set):") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 
```

Create Reproducible Brier Score: 
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}
```


## Logistic Regression: 
Brier's Score: 
```{r}
glm.BCW<-glm(malignant ~., data = BCW, subset = train.BCW, family = binomial)
BCW.probs<-predict(glm.BCW, newdata = BCW.test, type = "response") 
GLM_Brier = brier_score(V2.test, BCW.probs)
GLM_Brier
```
Confusion Matrix & Accuracy Score: 

Threshold: A threshold of .3 was chosen to ensure high sensitivity with respect to specify. A threshold of .3 helps reduce the risk of false negatives which would have devastating consequences with respect to classifying tumors.  

Confusion Matrix & Accuracy:
```{r}
glm.pred <- rep(0, nrow(BCW.test)) #size of the test set.
glm.pred[BCW.probs > 0.3] = 1#.3threshold for classification
table(glm.pred, V2.test)
GLM.Accuracy = mean(glm.pred == V2.test)
GLM.Accuracy
```
Roc Curve: 
```{r}
roc_obj.glm <- roc(V2.test, BCW.probs)
plot(roc_obj.glm, type = "s")
GLM.AUC = auc(roc_obj.glm)
GLM.AUC
```
Calibration CUrve: 
```{r}
glm.calplot= valProbggplot(BCW.probs, V2.test)
glm.calplot
```

## Stepwises: 
'both' allowed the model to evaluate predictors using both forward and backward steps. It's final model includes only 13 predictors.  
```{r, warning=FALSE}
set.seed(123)
glm.step <- step(glm.BCW, direction = "both", trace = FALSE) # Stepwise (both forward & backward) selection based on AIC. FALSE= dont see how got to choosen values. 
summary(glm.step)
```
Brier Score:
```{r}
step.probs<-predict(glm.step, newdata = BCW.test, type = "response") 
step_Brier = brier_score(V2.test, step.probs) 
step_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.step.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.step.pred[step.probs > 0.3] = 1 #.5threshold for classification
table(BCW.step.pred, V2.test)
Accuracy.step = mean(BCW.step.pred == V2.test)
Accuracy.step
```

Roc Curve: 
```{r}
roc_obj.step <- roc(V2.test, step.probs)
plot(roc_obj.step, type = "s")
step.AUC = auc(roc_obj.step)
step.AUC
```
Calibration CUrve: 
```{r}
step.calplot = valProbggplot(step.probs, V2.test)
step.calplot
```


## Ridge Regression: 
```{r}
set.seed(123)
x <- model.matrix(malignant ~ ., BCW)[, -1] #make matrixes
y <- BCW$malignant
ridge.mod<-glmnet(x[train.BCW,], y[train.BCW], alpha = 0, family = binomial)
cv.out <- cv.glmnet(x[train.BCW, ], y[train.BCW], alpha = 0, family = binomial)
plot(cv.out)
bestlam <- cv.out$lambda.min #May prefer 1-SE not minimum but here minimum. 
bestlam
```

Brier Score:
```{r}
ridge.probs <- predict(ridge.mod, s = bestlam,newx = x[!train.BCW, ], type = "response")
Ridge_Brier = brier_score(V2.test, ridge.probs)
Ridge_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.ridge.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.ridge.pred[ridge.probs > .3] <- 1 #.3threshold for classification
ridge.probs <- as.vector(ridge.probs)
table(BCW.ridge.pred, V2.test)
Accuracy.Ridge = mean(BCW.ridge.pred == V2.test)
Accuracy.Ridge
```
ROC Curve:
```{r}
roc_obj.ridge <- roc(V2.test, ridge.probs)
plot(roc_obj.ridge, type = "s")
Ridge.AUC = auc(roc_obj.ridge)
Ridge.AUC
```
Calibration Plot: 
```{r}
Ridge.calplot<-valProbggplot(ridge.probs, V2.test)
Ridge.calplot
```
