---
title: "Breast Cancer Wisconsin Dataset"
author: "Aine"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading Libraries: 
```{r, include=FALSE}
library(psych)
library(BART)
library(tree)
library(dcurves)
library(randomForest)
library(gtsummary)
library(broom.helpers)
library(e1071)
library(caret) 
library(ggcorrplot)
library(MASS) #For descrimminate analysis
library(pROC) #ROC curve
library(dplyr) 
library(knitr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(glmnet)#Lasso/ Ridge
library(gam)
library(boot)
library(pls)
library(gbm)
library(dbarts)
library(class)
library(CalibrationCurves) 
library(corrplot)
```

Loading Dataset:
```{r}
#Breast Cancer Wisconsin (Diagnostic)
BCWdata <- read.csv("C:/Users/aineg/OneDrive - TNS Capital/Aine Folder/HealthDataScience/Semester Two/Predictive Modelling and Statistical Learning/Predictive Assignment/wisconsin_dbc.csv")
```
Review Dataset: 
```{r}
BCW <- BCWdata %>%
  dplyr::select(-"ID") %>%                                    # removing the ID
  mutate(malignant = ifelse(Diagnosis == "M", 1, 0)) %>%  #numeric
  dplyr::select(-"Diagnosis")                                 
summary(BCW)
cor_mat <- cor(BCW[sapply(BCW, is.numeric)], use="pairwise.complete.obs")
cor_mat <- as.matrix(cor_mat)
ggcorrplot(cor_mat)

correlation = as.data.frame(as.table(cor_mat)) %>%
  mutate(Var1 = as.character(Var1), Var2 = as.character(Var2)) %>% #lets filter duplicates out 
  filter(Var1 < Var2) %>% #removes duplicates
  filter(abs(Freq) > 0.7) %>% 
  arrange(desc(abs(Freq))) #decending order

total_correlation <- nrow(correlation)
total_correlation
```

```{r}
correlation %>%
  head(10) %>%
  kable(digits = 2, caption = "Top 10 Correlations:") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 

  
```
Assessing for Extreme Values: 
```{r}
desc <- describe(BCW, fast = TRUE)
desc[ , c("min", "max", "mean", "median", "range")]
```


Split into Test/ Train:
```{r}
set.seed(321321) #reproducible
n<-nrow(BCW)
train.BCW<-runif(n)<0.7
BCW.test<-BCW[!train.BCW,]
V2.test<-BCW.test$malignant
V2.test.factor <- as.factor(BCW.test$malignant)

table(BCW$malignant[train.BCW]) %>%
  kable(digits = 2, caption = "Number of Events (training set):") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 

table(BCW$malignant[!train.BCW])%>%
  kable(digits = 2, caption = "Number of Events (test set):") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                position = "center") 
```

Create Reproducible Brier Score: 
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}
```


## Logistic Regression: 
Brier's Score: 
```{r}
glm.BCW<-glm(malignant ~., data = BCW, subset = train.BCW, family = binomial)
BCW.probs<-predict(glm.BCW, newdata = BCW.test, type = "response") 
GLM_Brier = brier_score(V2.test, BCW.probs)
GLM_Brier
```
Confusion Matrix & Accuracy Score: 

Threshold: A threshold of .3 was chosen to ensure high sensitivity with respect to specify. A threshold of .3 helps reduce the risk of false negatives which would have devastating consequences with respect to classifying tumors.  

Confusion Matrix & Accuracy:
```{r}
glm.pred <- rep(0, nrow(BCW.test)) #size of the test set.
glm.pred[BCW.probs > 0.3] = 1#.3threshold for classification
table(glm.pred, V2.test)
GLM.Accuracy = mean(glm.pred == V2.test)
GLM.Accuracy
```
Roc Curve: 
```{r}
roc_obj.glm <- roc(V2.test, BCW.probs)
plot(roc_obj.glm, type = "s")
GLM.AUC = auc(roc_obj.glm)
GLM.AUC
```
Calibration CUrve: 
```{r}
glm.calplot= valProbggplot(BCW.probs, V2.test)
glm.calplot
```

## Stepwises: 
'both' allowed the model to evaluate predictors using both forward and backward steps. It's final model includes only 13 predictors.  
```{r, warning=FALSE}
set.seed(123)
glm.step <- step(glm.BCW, direction = "both", trace = FALSE) # Stepwise (both forward & backward) selection based on AIC. FALSE= dont see how got to choosen values. 
summary(glm.step)
```
Brier Score:
```{r}
step.probs<-predict(glm.step, newdata = BCW.test, type = "response") 
step_Brier = brier_score(V2.test, step.probs) 
step_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.step.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.step.pred[step.probs > 0.3] = 1 #.5threshold for classification
table(BCW.step.pred, V2.test)
Accuracy.step = mean(BCW.step.pred == V2.test)
Accuracy.step
```

Roc Curve: 
```{r}
roc_obj.step <- roc(V2.test, step.probs)
plot(roc_obj.step, type = "s")
step.AUC = auc(roc_obj.step)
step.AUC
```
Calibration CUrve: 
```{r}
step.calplot = valProbggplot(step.probs, V2.test)
step.calplot
```


## Ridge Regression: 
```{r}
set.seed(123)
x <- model.matrix(malignant ~ ., BCW)[, -1] #make matrixes
y <- BCW$malignant
ridge.mod<-glmnet(x[train.BCW,], y[train.BCW], alpha = 0, family = binomial)
cv.out <- cv.glmnet(x[train.BCW, ], y[train.BCW], alpha = 0, family = binomial)
plot(cv.out)
bestlam <- cv.out$lambda.min #May prefer 1-SE not minimum but here minimum. 
bestlam
```

Brier Score:
```{r}
ridge.probs <- predict(ridge.mod, s = bestlam,newx = x[!train.BCW, ], type = "response")
Ridge_Brier = brier_score(V2.test, ridge.probs)
Ridge_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.ridge.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.ridge.pred[ridge.probs > .3] <- 1 #.3threshold for classification
ridge.probs <- as.vector(ridge.probs)
table(BCW.ridge.pred, V2.test)
Accuracy.Ridge = mean(BCW.ridge.pred == V2.test)
Accuracy.Ridge
```
ROC Curve:
```{r}
roc_obj.ridge <- roc(V2.test, ridge.probs)
plot(roc_obj.ridge, type = "s")
Ridge.AUC = auc(roc_obj.ridge)
Ridge.AUC
```
Calibration Plot: 
```{r}
Ridge.calplot<-valProbggplot(ridge.probs, V2.test)
Ridge.calplot
```
## Lasso Model: 
```{r}
set.seed(321321)
lasso.mod<-glmnet(x[train.BCW,], y[train.BCW], alpha = 1, family = binomial)
cv.out <- cv.glmnet(x[train.BCW, ], y[train.BCW], alpha = 1, family = binomial)
plot(cv.out)
bestlam <- cv.out$lambda.min #May prefer 1-SE not minimum but here minimum.
bestlam
#cv.out$lambda.1se
Kept <- coef(cv.out, s = "lambda.min")
Kept[Kept[,1] != 0, ]
```
Brier's Score:
```{r}
lasso.probs <- predict(lasso.mod, s = bestlam,newx = x[!train.BCW, ], type = "response")
lasso_Brier = brier_score(V2.test, lasso.probs)#Ensure factors are 1 & 0. 
lasso_Brier
```
Confusion Matrix & Accuracy Score: 
```{r}
BCW.lasso.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.lasso.pred[lasso.probs > .3] <- 1 #.3threshold for classification
table(BCW.lasso.pred, V2.test)
Accuracy.lasso = mean(BCW.lasso.pred == V2.test)
Accuracy.lasso
```
Roc Curve:
```{r}
roc_obj.lasso <- roc(V2.test, as.vector(lasso.probs))
plot(roc_obj.lasso, type = "s")
lasso.AUC = auc(roc_obj.lasso)
lasso.AUC
```
Calibration Plot: 
```{r}
lasso.calplot<-valProbggplot(lasso.probs, V2.test)
lasso.calplot
```
Matrix Test/ Train: 
```{r}
set.seed(123)
standardized.X <- scale(BCW[, names(BCW) != "malignant"]) #standardize & sure malignant isnt 
Y <- as.factor(BCW$malignant)#y must be a factor in PCA
X <- standardized.X[train.BCW, ]
XTest <- standardized.X[!train.BCW, ]
Ytrain <- Y[train.BCW]
Ytest <- Y[!train.BCW]
Ytest_numeric = as.numeric(as.character(Ytest)) #going factor-numeric causes it to fail
Ytrain_numeric = as.numeric(as.character(Ytrain))
```

## PLS-DA Approach: 
```{r}
plsda.fit <- plsda(X, Ytrain, cv = TRUE)
validationplot(plsda.fit, val.type = "MSEP")
```
Brier's Score
```{r}
plsda.fit <- plsda(X, Ytrain, ncomp = 1)
plsda.probs <- predict(plsda.fit, newdata = XTest, type = "prob")

prob_M <- plsda.probs[, "1", 1]
Ytest_numeric <- as.numeric(as.character(Ytest))#becuase made y a factor back to numeric

plsda_Brier <- brier_score(Ytest_numeric, prob_M)
plsda_Brier
```
Confusion Matrix & Accuracy Score:
```{r}
BCW.plsda.pred <- rep(0, length(prob_M))
BCW.plsda.pred[prob_M > 0.3] <- 1
table(BCW.plsda.pred, Ytest)
Accuracy.plsda = mean(BCW.plsda.pred == Ytest)
Accuracy.plsda
```

Roc Curve: 
```{r}
roc_obj.plsda <- roc(Ytest_numeric, prob_M)
plot(roc_obj.plsda, type = "s")
plsda.AUC = auc(roc_obj.plsda)
plsda.AUC
```
Calibration Plot: 
```{r}
plsda.calplot<-valProbggplot(prob_M, Ytest_numeric)
plsda.calplot
```
## Assessment: 
```{r}
results <- data.frame(
  Model = c("Logistic Regression", "Best Subset", "Ridge Model", "Lasso Model", 'PLS-DA Model'),  Accuracy = round(c(GLM.Accuracy,Accuracy.step, Accuracy.Ridge, Accuracy.lasso, Accuracy.plsda), 3),
  Brier = round(c(GLM_Brier, step_Brier, Ridge_Brier, lasso_Brier, plsda_Brier), 3),
  AUC = round(c(GLM.AUC,step.AUC, Ridge.AUC, lasso.AUC, plsda.AUC), 3)
)
#To get the highlighted colours: %in% -this is important 
best_accuracy <- which(results$Accuracy == max(results$Accuracy))
best_brier    <- which(results$Brier == min(results$Brier))
best_auc      <- which(results$AUC == max(results$AUC))


results %>%
  kable(caption = "Accuracy, Brier & AUC") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, position = "center") %>%
  column_spec(2, 
              bold = 1:nrow(results) %in% best_accuracy,
              background = ifelse(1:nrow(results) %in% best_accuracy, 
                                  "#c6efce", "white")) %>%
  column_spec(3, 
              bold = 1:nrow(results) %in% best_brier,
              background = ifelse(1:nrow(results) %in% best_brier, 
                                  "#c6efce", "white")) %>%
  column_spec(4, 
              bold = 1:nrow(results) %in% best_auc,
              background = ifelse(1:nrow(results) %in% best_auc, 
                                  "#c6efce", "white"))
```
Roc Curves: 
```{r}
plot(roc_obj.glm, type = "s", col = "blue",main = "ROC Curves — Classification Models",xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc_obj.step, type = "s", col = "maroon") 
lines(roc_obj.ridge, type = "s", col = "purple")
lines(roc_obj.lasso, type = "s", col = "navy")
lines(roc_obj.plsda, type = "s", col = "goldenrod")

legend("bottomright",
       legend = c("Logistic", "Stepwise", "Ridge", "Lasso", "PLS-DA"),
       col = c("blue", "maroon", "purple", "navy", "goldenrod"),
       lwd = 2,cex = 0.8)
```


Calibration Curves & Info: 
```{r}

grid.arrange(
  glm.calplot$ggPlot + ggtitle("Logistic Regression"),
  step.calplot$ggPlot + ggtitle("Best Subset"),
  Ridge.calplot$ggPlot + ggtitle("Ridge Model"),
  lasso.calplot$ggPlot + ggtitle("Lasso Model"),
  plsda.calplot$ggPlot + ggtitle('PLS-DA Model'),
  ncol = 3
)

cal.results <- data.frame(
  Model = c("Logistic Regression", "Best Subset", "Ridge Model", "Lasso Model", 'PLS-DA Model'),
  Cal.Intercept = round(c(
    glm.calplot$stats["Intercept"],
    step.calplot$stats["Intercept"],
    Ridge.calplot$stats["Intercept"],
    lasso.calplot$stats["Intercept"],
    plsda.calplot$stats["Intercept"]), 3),
  Cal.Slope = round(c(
    glm.calplot$stats["Slope"],
    step.calplot$stats["Slope"],
    Ridge.calplot$stats["Slope"],
    lasso.calplot$stats["Slope"],
    plsda.calplot$stats["Slope"]), 3)
)
best_intercept <- which.min(abs(cal.results$Cal.Intercept - 0))
best_slope <- which.min(abs(cal.results$Cal.Slope - 1)) 

cal.results %>%
  kable(caption = "Calibration")%>%
   column_spec(2, bold = ifelse(1:nrow(cal.results) == best_intercept, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_intercept, "#c6efce", "white")) %>%
  column_spec(3, bold = ifelse(1:nrow(cal.results) == best_slope, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_slope, "#c6efce", "white")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE,      position = "center")
```


## Linear Discriminant Analysis:
```{r}
lda.fit <- lda(malignant ~., data = BCW, subset = train.BCW)
lda.pred <- predict(lda.fit, BCW.test)
lda.class <- lda.pred$class
lda.probs <- lda.pred$posterior
prob_M <- lda.probs[, 2]
```
Brier's Score: 
```{r}
lda_Brier <- brier_score(V2.test, prob_M)
lda_Brier
```

Confusion Matrix & Accuracy Score:
```{r}
BCW.lda.pred <- rep(0, nrow(BCW.test)) #size of the test set.
BCW.lda.pred[prob_M > .3] <- 1 
table(BCW.lda.pred, V2.test)
Accuracy.lda = mean(BCW.lda.pred == V2.test)
Accuracy.lda
```

Roc Curve:
```{r}
roc_obj.lda <- roc(V2.test, prob_M)
plot(roc_obj.lda, type = "s")
lda.AUC = auc(roc_obj.lda)
lda.AUC
```
Calibration Plot: 
```{r}
lda.calplot<-valProbggplot(prob_M, V2.test)
lda.calplot
```
## Naive Bayes
```{r}
nb.fit <- naiveBayes(as.factor(malignant) ~ ., data = BCW, subset = train.BCW) #TEST/ TRAIN SPLIT
class(BCW$malignant)
```
Brier's Score:
```{r}
nb.pred <- predict(nb.fit, BCW.test, type = "raw")  #again predicted on the test set 
nb.probs <- nb.pred[, 2]#extracts column of probability=malignant 
NB_Brier = brier_score(V2.test, nb.probs)#Ensure factors are 1 & 0. 
NB_Brier
colnames(nb.pred)

```

Confusion Matrix & Accuracy: 
```{r}
nb.class= rep(0, nrow(BCW.test))  
nb.class[nb.probs > 0.3] = 1  
nb.class <- factor(nb.class, levels = c(0,1))   
table(nb.class, V2.test)
NB.Accuracy <- mean(nb.class == V2.test)
NB.Accuracy
```

Roc Curve:
```{r}
roc_obj.NB <- roc(V2.test, nb.probs)
plot(roc_obj.NB, type = "s", col = "blue")

NB.AUC = auc(roc_obj.NB)
NB.AUC
```

Calibration Curve: 
```{r}
NB.calplot = valProbggplot(nb.probs, V2.test)
NB.calplot
```
## K-NN
```{r}
set.seed(123)
k.values <- 1:20 #k can be 1 to 20
cv.error <- rep(NA, 20) #stores misclassification error
for(i in k.values){
  pred <- knn.cv(train = X,cl = Ytrain,k = i) #leave-one-out CV 
  cv.error[i] <- mean(pred != Ytrain) #stores misclassification erro
}
best.k <- which.min(cv.error)
best.k
```
Brier Score: 
```{r}
knn.fit <- knn(X, XTest, Ytrain, k = best.k, prob = TRUE)
knn.probs <- attr(knn.fit, "prob") #knn for classification couldnt use predict? 
prob_M <- ifelse(knn.fit == 1, knn.probs, 1 - knn.probs)
knn_Brier = brier_score(Ytest_numeric, prob_M)
knn_Brier
```
Accuracy & Confusion Matrix:
```{r}
knn.class= rep(0, nrow(BCW.test))  
knn.class[prob_M > 0.3] = 1  
table(knn.class, Ytest_numeric)
knn.Accuracy = mean(knn.class==Ytest_numeric)
knn.Accuracy
```
Roc Curve:
```{r}
roc_obj.knn<- roc(Ytest_numeric, prob_M)
plot(roc_obj.knn, type = 's')
knn.AUC = auc(roc_obj.knn)
knn.AUC
```
Calibration Curve: 
```{r}
knn.calplot<-valProbggplot(prob_M, Ytest_numeric)
knn.calplot 
```

## Assessment: 
```{r}
results <- data.frame(
  Model = c("Linear Discriminat Analysis", 'Naive Bayes', 'K-NN'),  Accuracy = round(c(Accuracy.lda,NB.Accuracy, knn.Accuracy), 3),
  Brier = round(c(lda_Brier, NB_Brier, knn_Brier), 3),
  AUC = round(c(lda.AUC,NB.AUC, knn.AUC), 3)
)
#To get the highlighted colours: %in% -this is important 
best_accuracy <- which(results$Accuracy == max(results$Accuracy))
best_brier    <- which(results$Brier == min(results$Brier))
best_auc      <- which(results$AUC == max(results$AUC))


results %>%
  kable(caption = "Accuracy, Brier & AUC") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, position = "center") %>%
  column_spec(2, 
              bold = 1:nrow(results) %in% best_accuracy,
              background = ifelse(1:nrow(results) %in% best_accuracy, 
                                  "#c6efce", "white")) %>%
  column_spec(3, 
              bold = 1:nrow(results) %in% best_brier,
              background = ifelse(1:nrow(results) %in% best_brier, 
                                  "#c6efce", "white")) %>%
  column_spec(4, 
              bold = 1:nrow(results) %in% best_auc,
              background = ifelse(1:nrow(results) %in% best_auc, 
                                  "#c6efce", "white"))
```
Roc Curves: 
```{r}
plot(roc_obj.lda, type = "s", col = "blue",main = "ROC Curves — Classification Models",xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc_obj.NB, type = "s", col = "maroon") 
lines(roc_obj.knn, type = "s", col = "purple")


legend("bottomright",
       legend = c("Linear Discriminat Analysis", 'Naive Bayes', 'K-NN'),
       col = c("blue", "maroon", "purple"),
       lwd = 2,cex = 0.8)
```


Calibration Curves & Info: 
```{r}

grid.arrange(
  lda.calplot$ggPlot + ggtitle("Linear Discriminat Analysis"),
  NB.calplot$ggPlot + ggtitle("Naive Bayes"),
  knn.calplot$ggPlot + ggtitle("K-NN"),
  ncol = 3
)

cal.results <- data.frame(
  Model = c("Linear Discriminat Analysis", 'Naive Bayes', 'K-NN'),
  Cal.Intercept = round(c(
    lda.calplot$stats["Intercept"],
    NB.calplot$stats["Intercept"],
    knn.calplot$stats["Intercept"]), 3),
  Cal.Slope = round(c(
    lda.calplot$stats["Slope"],
    NB.calplot$stats["Slope"],
    knn.calplot$stats["Slope"]), 3)
)
best_intercept <- which.min(abs(cal.results$Cal.Intercept - 0))
best_slope <- which.min(abs(cal.results$Cal.Slope - 1)) 

cal.results %>%
  kable(caption = "Calibration")%>%
   column_spec(2, bold = ifelse(1:nrow(cal.results) == best_intercept, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_intercept, "#c6efce", "white")) %>%
  column_spec(3, bold = ifelse(1:nrow(cal.results) == best_slope, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_slope, "#c6efce", "white")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE,      position = "center")
```
## Tree
```{r}
set.seed(555)
tree.BCW <- tree(as.factor(malignant)~ . , BCW, subset =train.BCW)
cv.BCW <- cv.tree(tree.BCW, FUN = prune.misclass)
best <- cv.BCW$size[which.min(cv.BCW$dev)]
best
prune.fit<- prune.misclass(tree.BCW, best = best)
summary(prune.fit)
plot(prune.fit)
text(prune.fit, pretty = 0)
```
Brier Score:
```{r}
prune.probs <- predict(prune.fit, newdata = BCW[!train.BCW, ], type = "vector")
prob_M <- prune.probs[, 2]
prune_Brier <- brier_score(V2.test, prob_M)
prune_Brier 
```
Accuracy & Confusion Matrix: 
```{r}
prune.class <- predict(prune.fit, BCW[!train.BCW, ],type = "class")
table(prune.class, V2.test)
tree.Accuracy <- mean(prune.class == V2.test.factor)
tree.Accuracy
```
ROC Curve: 
```{r}
roc_obj.tree<- roc(V2.test, prob_M)
plot(roc_obj.tree, type = 's')
tree.AUC = auc(roc_obj.tree)
tree.AUC
```
Calibration Plot: 
```{r}
tree.calplot<-valProbggplot(prob_M, V2.test)
tree.calplot
```

## Random Forest: 
CV for m: 
```{r}
set.seed(321)
mtry.values <- c(2, 5, 8, 10, 15)

mtry.errors <- sapply(mtry.values, function(m){
  fit <- randomForest(as.factor(malignant) ~., data = BCW[train.BCW, ], mtry = m, ntree = 500)
  return(fit$err.rate[500, "OOB"])  
})

best.mtry <- mtry.values[which.min(mtry.errors)]
best.mtry
rf.fit <- randomForest(as.factor(malignant) ~ ., data = BCW, subset =train.BCW, mtry = best.mtry, importance = TRUE)
```
Brier Score: 
```{r}
rf.probs <- predict(rf.fit, newdata = BCW.test, type = "prob")
prob_M <- rf.probs[, 2]
rf_Brier <- brier_score(V2.test, prob_M)
rf_Brier
```
Accuracy & Confusion Matrix:
```{r}
rf.class= rep(0, nrow(BCW.test))  
rf.class[prob_M > 0.3] = 1
rf.class = factor(rf.class, levels = c(0,1))
table(rf.class, V2.test)
rf.Accuracy = mean(rf.class==V2.test)
rf.Accuracy
```

Roc Curve:
```{r}
roc_obj.rf<- roc(V2.test, prob_M)
plot(roc_obj.rf, type = 's')
rf.AUC = auc(roc_obj.rf)
rf.AUC
```
Calibration Curve: 
```{r}
rf.calplot<-valProbggplot(prob_M, V2.test)
rf.calplot 
```
## Boosting:
I trialed different Boosting Models:
with different number of trees, different shrinkage and different interaction depths. 
```{r}
set.seed(321)
boost.fit1000 <- gbm(malignant ~., data = BCW[train.BCW, ],distribution = "bernoulli",n.trees = 1000,interaction.depth = 4,shrinkage = 0.01,cv.folds = 5)

best.trees <- gbm.perf(boost.fit1000, method = "cv")
best.trees
```
Shrinkage:
```{r}
set.seed(321)
shrinkage.values <- c(0.001, 0.01, 0.05, 0.1, 0.2) 

shrinkage.errors <- sapply(shrinkage.values, function(s){
  fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = 585,    interaction.depth = 4,shrinkage = s, cv.folds = 5) 
  return(min(fit$cv.error))       
})

best.shrinkage <- shrinkage.values[which.min(shrinkage.errors)]
best.shrinkage
```
Interaction Depth: 
```{r}
set.seed(321)
id <- c(2, 3, 4, 5) 

depth.errors <- sapply(id, function(d){
  fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = 585,    interaction.depth = d,shrinkage = 0.2, cv.folds = 5)
  return(min(fit$cv.error))
})

best.depth <- id[which.min(depth.errors)]
best.depth
```
Brier Score: 
```{r}
boost.fit <- gbm(malignant ~ ., data = BCW[train.BCW, ], distribution = "bernoulli", n.trees = best.trees,    interaction.depth = best.depth,shrinkage = best.shrinkage) #optimum fit

boost.probs <- predict(boost.fit, newdata = BCW.test, type = "response") #vector of probs 
boost_Brier <- brier_score(V2.test, boost.probs)
boost_Brier
```
Accuracy & Confusion Matrix:
```{r}
boost.class= rep(0, nrow(BCW.test))  
boost.class[boost.probs > 0.3] = 1  
boost.class = factor(boost.class, levels = c(0, 1))
table(boost.class, V2.test)
boost.Accuracy = mean(boost.class==V2.test)
boost.Accuracy
```

Roc Curve:
```{r}
roc_obj.boost<- roc(V2.test, boost.probs)
plot(roc_obj.boost, type = 's')
boost.AUC = auc(roc_obj.boost)
boost.AUC
```
Calibration Curve: 
```{r}
boost.calplot<-valProbggplot(boost.probs, V2.test)
boost.calplot 
```
## Bayesian Additive Regression Trees
```{r}
set.seed(555)
xbart_result <- xbart(X, Ytrain_numeric, n.trees = c(50, 75, 100, 200), method = "k-fold", n.reps = 10)
best.ntree <- c(50, 75, 100, 200)[which.min(colMeans(xbart_result))]
best.ntree
```
Fit:
```{r}
set.seed(321)
bart.fit <- bart(X, Ytrain_numeric,x.test = XTest,ntree = best.ntree,ndpost = 500,nskip = 100,keeptrees = FALSE, verbose = FALSE)
bart.fit2 <- bart(X, Ytrain_numeric,x.test = XTest,ntree = best.ntree,ndpost = 1000,nskip = 500,keeptrees = FALSE, verbose = FALSE)
bart.fit3 <- bart(X, Ytrain_numeric,x.test = XTest,ntree = best.ntree,ndpost = 2000,nskip = 1000,keeptrees = FALSE, verbose = FALSE)
# yhat.test is a matrix of draws — need to take mean
prob1 <- pnorm(colMeans(bart.fit$yhat.test))
prob2 <- pnorm(colMeans(bart.fit2$yhat.test))
prob3 <- pnorm(colMeans(bart.fit3$yhat.test))
Brier1 <- brier_score(Ytest_numeric, prob1)
Brier2 <- brier_score(Ytest_numeric, prob2)
Brier3 <- brier_score(Ytest_numeric, prob3)
Brier1
Brier2
Brier3
bart_Brier =Brier2
prob_M = prob2
```
Accuracy & Confusion Matrix:
```{r}
bart.class <- factor(ifelse(prob_M > 0.5, 1, 0), levels = c(0, 1))
table(bart.class, V2.test)
Accuracy.bart <- mean(bart.class == V2.test.factor)
Accuracy.bart
```

Roc Curve:
```{r}
roc_obj.bart<- roc(V2.test, prob_M)
plot(roc_obj.bart, type = 's')
bart.AUC = auc(roc_obj.bart)
bart.AUC
```
Calibration Curve: 
```{r}
bart.calplot<-valProbggplot(prob_M, V2.test)
bart.calplot 
```
## Assessment: 
```{r}
results <- data.frame(
  Model = c("Tree", "Random Forest", 'Boosting', 'BART'),  Accuracy = round(c(tree.Accuracy, rf.Accuracy,boost.Accuracy, Accuracy.bart), 3),
  Brier = round(c(prune_Brier, rf_Brier, boost_Brier, bart_Brier), 3),
  AUC = round(c(tree.AUC, rf.AUC,boost.AUC, bart.AUC), 3)
)
#To get the highlighted colours: %in% -this is important 
best_accuracy <- which(results$Accuracy == max(results$Accuracy))
best_brier    <- which(results$Brier == min(results$Brier))
best_auc      <- which(results$AUC == max(results$AUC))


results %>%
  kable(caption = "Accuracy, Brier & AUC") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, position = "center") %>%
  column_spec(2, 
              bold = 1:nrow(results) %in% best_accuracy,
              background = ifelse(1:nrow(results) %in% best_accuracy, 
                                  "#c6efce", "white")) %>%
  column_spec(3, 
              bold = 1:nrow(results) %in% best_brier,
              background = ifelse(1:nrow(results) %in% best_brier, 
                                  "#c6efce", "white")) %>%
  column_spec(4, 
              bold = 1:nrow(results) %in% best_auc,
              background = ifelse(1:nrow(results) %in% best_auc, 
                                  "#c6efce", "white"))
```
ROC Curves: 
```{r}
plot(roc_obj.tree, type = "s", col = "blue",main = "ROC Curves — Classification Models",xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc_obj.rf, type = "s", col = "gold") 
lines(roc_obj.boost, type = "s", col = "maroon") 
lines(roc_obj.bart, type = "s", col = "purple")


legend("bottomright",
       legend = c("Tree","Random Forest", 'Boosting', 'BART'),
       col = c("blue","gold", "maroon", "purple"),
       lwd = 2,cex = 0.8)
```
Calibration Curves: 
```{r}
grid.arrange(
  tree.calplot$ggPlot + ggtitle("Tree"),
  rf.calplot$ggPlot + ggtitle("Random Forest"),
  boost.calplot$ggPlot + ggtitle('Boosting'),
  bart.calplot$ggPlot + ggtitle('BART'),
  ncol = 3
)

cal.results <- data.frame(
  Model = c("Tree", "Random Forest", 'Boosting', 'BART'),
  Cal.Intercept = round(c(
    tree.calplot$stats["Intercept"],
    rf.calplot$stats["Intercept"],
    boost.calplot$stats["Intercept"],
    bart.calplot$stats["Intercept"]), 3),
  Cal.Slope = round(c(
    tree.calplot$stats["Slope"],
    rf.calplot$stats["Slope"],
    boost.calplot$stats["Slope"],
    bart.calplot$stats["Slope"]), 3)
)
best_intercept <- which.min(abs(cal.results$Cal.Intercept - 0))
best_slope <- which.min(abs(cal.results$Cal.Slope - 1)) 

cal.results %>%
  kable(caption = "Calibration")%>%
   column_spec(2, bold = ifelse(1:nrow(cal.results) == best_intercept, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_intercept, "#c6efce", "white")) %>%
  column_spec(3, bold = ifelse(1:nrow(cal.results) == best_slope, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_slope, "#c6efce", "white")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE,      position = "center")
```


## Conclusion:
```{r}
results <- data.frame(
  Model = c("Lasso", "LDA", 'Boosting'),  Accuracy = round(c(Accuracy.lasso, Accuracy.lda,boost.Accuracy), 3),
  Brier = round(c(lasso_Brier, lda_Brier, boost_Brier), 3),
  AUC = round(c(lasso.AUC, lda.AUC,boost.AUC), 3)
)
#To get the highlighted colours: %in% -this is important 
best_accuracy <- which(results$Accuracy == max(results$Accuracy))
best_brier    <- which(results$Brier == min(results$Brier))
best_auc      <- which(results$AUC == max(results$AUC))


results %>%
  kable(caption = "Accuracy, Brier & AUC") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, position = "center") %>%
  column_spec(2, 
              bold = 1:nrow(results) %in% best_accuracy,
              background = ifelse(1:nrow(results) %in% best_accuracy, 
                                  "#c6efce", "white")) %>%
  column_spec(3, 
              bold = 1:nrow(results) %in% best_brier,
              background = ifelse(1:nrow(results) %in% best_brier, 
                                  "#c6efce", "white")) %>%
  column_spec(4, 
              bold = 1:nrow(results) %in% best_auc,
              background = ifelse(1:nrow(results) %in% best_auc, 
                                  "#c6efce", "white"))
```
ROC Curves: 
```{r}
plot(roc_obj.lasso, type = "s", col = "blue",main = "ROC Curves — Classification Models",xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc_obj.lda, type = "s", col = "gold") 
lines(roc_obj.boost, type = "s", col = "maroon") 


legend("bottomright",
       legend = c("Lasso","LDA", 'Boosting'),
       col = c("blue","gold", "maroon"),
       lwd = 2,cex = 0.8)
```
Calibration Curves: 
```{r}
grid.arrange(
  lasso.calplot$ggPlot + ggtitle("Lasso"),
  lda.calplot$ggPlot + ggtitle("LDA"),
  boost.calplot$ggPlot + ggtitle('Boosting'),
  ncol = 3
)

cal.results <- data.frame(
  Model = c("Lasso", "LDA", 'Boosting'),
  Cal.Intercept = round(c(
    lasso.calplot$stats["Intercept"],
    lda.calplot$stats["Intercept"],
    boost.calplot$stats["Intercept"]), 3),
  Cal.Slope = round(c(
    lasso.calplot$stats["Slope"],
    lda.calplot$stats["Slope"],
    boost.calplot$stats["Slope"]), 3)
)
best_intercept <- which.min(abs(cal.results$Cal.Intercept - 0))
best_slope <- which.min(abs(cal.results$Cal.Slope - 1)) 

cal.results %>%
  kable(caption = "Calibration")%>%
   column_spec(2, bold = ifelse(1:nrow(cal.results) == best_intercept, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_intercept, "#c6efce", "white")) %>%
  column_spec(3, bold = ifelse(1:nrow(cal.results) == best_slope, TRUE, FALSE),
              background = ifelse(1:nrow(cal.results) == best_slope, "#c6efce", "white")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE,      position = "center")
```

## Best Predictors: 
```{r}
boost.importance <- summary(boost.fit, plotit = FALSE)

ggplot(boost.importance[1:10, ],  
       aes(x = reorder(var, rel.inf), y = rel.inf)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Boosting Predictors",
       x = "Variable",
       y = "Relative Influence (%)") +
  theme_minimal()

Kept <- coef(cv.out, s = "lambda.min")
predictors = Kept[Kept[,1] != 0, ] 

predictors %>%
  kable(caption = "Predictors Kept")%>%
    kable_styling(bootstrap_options = c("striped", "hover"),full_width = FALSE,position = "center")
```

Clinical Usefullness:

```{r}
df_dca <- data.frame(
  V2 = V2.test,
  boost = as.vector(boost.probs)
)

dcurves::dca(
  V2 ~ boost,
  data = df_dca,
  thresholds = seq(0, 1, 0.01),
  label = list(BCWpred = "Boosting Model")
)

dca_results <- dcurves::dca(
  V2 ~ boost,
  data = df_dca,
  as_probability = "boost",
  thresholds = c(0.05, 0.35, 0.75)
)

dca_results %>%
  as_tibble() %>%
  select(label, threshold, net_benefit) %>%
  gt::gt() %>%
  gt::fmt_percent(columns = threshold, decimals = 0) %>%
  gt::fmt(columns = net_benefit, fns = function(x) style_sigfig(x, digits = 3)) %>%
  gt::cols_label(
    label = "Strategy",
    threshold = "Decision Threshold",
    net_benefit = "Net Benefit"
  ) %>%
  gt::cols_align("left", columns = label)
```

